\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amsthm}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{dsfont}  % \mathds{1} indicator (OldStandard-Math lacks a bold digit 1)
% Old-style journal look (after tex.stackexchange.com/q/645339): OldStandard text + matching
% OpenType math, small-caps headings, run-in paragraphs. Requires lualatex (pinned in .latexmkrc).
\usepackage{OldStandard}
\usepackage{unicode-math}
\setmathfont{OldStandard-Math.otf}
\usepackage{titlesec}
\usepackage{fancyvrb}
\usepackage[hidelinks]{hyperref}

\linespread{1.15}
\setlength{\parindent}{2em}
\titleformat{\section}{\normalfont\large\scshape\bfseries}{\thesection.}{0.6em}{}
\titleformat{\subsection}{\normalfont\scshape\bfseries}{\thesubsection.}{0.5em}{}
\titleformat{\paragraph}[runin]{\bfseries\scshape}{}{0pt}{}[.]
\titlespacing*{\paragraph}{\parindent}{0.4em}{0.6em}

\newtheoremstyle{oldplain}{}{}{\itshape}{}{\scshape}{.}{.5em}{}
\newtheoremstyle{oldrem}{}{}{\normalfont}{}{\scshape}{.}{.5em}{}
\theoremstyle{oldplain}
\newtheorem{proposition}{Proposition}
\theoremstyle{oldrem}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\newcommand{\Pgen}{P_{\mathrm{gen}}}
\newcommand{\ins}{\mathrm{ins}}
\newcommand{\del}{\mathrm{del}}
\newcommand{\indic}{\mathds{1}}

\title{\scshape The V(D)J recombination model and fast generation probabilities in \texttt{vdjtools}}
\author{Mikhail Shugay\textsuperscript{1,2,3}\thanks{Correspondence:
  \href{mailto:mikhail.shugay@gmail.com}{mikhail.shugay@gmail.com}}}
\date{%
  \small
  \textsuperscript{1}Institute of Translational Medicine, Russian National Medical State University, Moscow, Russia\\
  \textsuperscript{2}Department of Genomics of Adaptive Immunity, Shemyakin--Ovchinnikov Institute of Bioorganic Chemistry RAS, Moscow, Russia\\
  \textsuperscript{3}Institute of Molecular Biology NAS RA, Erevan, Armenia\\[6pt]
  \normalsize\textit{Appendix M --- \texttt{vdjtools} model engine}%
}

\renewcommand{\thesection}{M.\arabic{section}}

\begin{document}
\maketitle

\begin{abstract}
\noindent
The \texttt{vdjtools} model engine reimplements the Murugan--Marcou generative model of V(D)J
recombination \cite{Murugan2012,Marcou2018} on the AIRR schema \cite{VanderHeiden2018}, storing every
model factor as a tidy \texttt{polars} table so that the Bayesian-network structure is declared as data
rather than baked into a file format. On top of this representation it computes the generation
probability ($\Pgen$) of a CDR3 --- the total probability that the stochastic recombination process
produces a given sequence --- both for nucleotide queries (an exact scenario sum) and, more delicately,
for amino-acid queries, where one must marginalize over all synonymous nucleotide realizations. For the
amino-acid case we use the transfer-matrix factorization of \cite{Murugan2012,Sethna2019}, but re-derive
it from a single structural observation --- \emph{that the only coupling across a recombination-block
boundary is codon translation, never the insertion Markov chain} --- which yields a clean left/right
partial-sum split (\S\ref{sec:aa}) that our native C++ core evaluates faster than \texttt{OLGA}
(\S\ref{sec:bench}) while matching it to machine precision. The same tables and scenario enumeration
drive sequence generation (ancestral sampling) and model inference (expectation--maximization), which we
summarize in \S\ref{sec:em}.
\end{abstract}

\section{The generative recombination model}\label{sec:model}

A rearranged receptor chain is the output of a stochastic process that (i) selects germline segments,
(ii) trims their ends by exonuclease deletion, and (iii) inserts non-templated nucleotides at the
junctions. Following \cite{Murugan2012,Marcou2018} we treat a \emph{recombination scenario} $r$ --- the
full list of hidden events --- as a sample from a Bayesian network whose factors are the model
parameters. For a $VDJ$ locus (TRB, TRD, IGH) a scenario is
\begin{equation}\label{eq:scenario}
  r=\bigl(V,\,J,\,D,\;\del_V,\,\del_J,\,\del_D^{5'},\,\del_D^{3'},\;\mathbf{m}^{VD},\,\mathbf{m}^{DJ}\bigr),
\end{equation}
with probability
\begin{equation}\label{eq:pvdj}
  P(r)=P(V)\,P(J)\,P(D\mid J)\;
       P(\del_V\!\mid V)\,P(\del_J\!\mid J)\,P(\del_D^{5'},\del_D^{3'}\!\mid D)\;
       P_{\ins}(\mathbf{m}^{VD})\,P_{\ins}(\mathbf{m}^{DJ}).
\end{equation}
A $VJ$ locus (TRA, TRG, IGK, IGL) has no $D$ segment; its scenario drops $D$, the two $D$-deletions and
one insertion, and the gene factor becomes the joint $P(V)\,P(J\mid V)$ with a single $VJ$ insertion.
The observed CDR3 nucleotide sequence $\sigma(r)$ is the concatenation, from the conserved Cys\,104 to
the conserved Phe/Trp\,118 inclusive,
\begin{equation}\label{eq:concat}
  \sigma(r)=\underbrace{\hat V}_{\text{V germline}}\;
            \underbrace{\mathbf{m}^{VD}}_{\text{N1}}\;
            \underbrace{\hat D}_{\text{D germline}}\;
            \underbrace{\mathbf{m}^{DJ}}_{\text{N2}}\;
            \underbrace{\hat J}_{\text{J germline}},
\end{equation}
where $\hat V,\hat D,\hat J$ are the germline segments after the indicated deletions.

\paragraph{Insertion chains} Each non-templated segment $\mathbf{m}=m_1\cdots m_\ell$ is drawn from a
length distribution and a first-order Markov chain:
\begin{equation}\label{eq:ins}
  P_{\ins}(\mathbf{m}) = P_{\ins}(\ell)\; b(m_1)\prod_{i=2}^{\ell} R(m_i\mid m_{i-1}),
\end{equation}
with $b$ the first-nucleotide (stationary) bias and $R$ the $4\times4$ dinucleotide transition matrix.
Following IGoR/\texttt{OLGA}, the $VD$ chain is read $5'\!\to\!3'$ and the $DJ$ chain $3'\!\to\!5'$, so
$b_{DJ}$ biases the $J$-proximal insertion nucleotide.

\paragraph{Generation probability} The quantity of interest is the probability of an \emph{observed}
CDR3, obtained by summing over every scenario that could have produced it,
\begin{equation}\label{eq:pgen}
  \Pgen(\sigma) \;=\; \sum_{r\,:\,\sigma(r)=\sigma} P(r),
  \qquad
  \Pgen(\alpha) \;=\; \sum_{\sigma\,:\,\tau(\sigma)=\alpha}\Pgen(\sigma),
\end{equation}
where $\tau$ is translation: an amino-acid query $\alpha$ additionally marginalizes over all synonymous
nucleotide sequences. Both sums are finite but astronomically large if taken literally; \S\ref{sec:nt}
and \S\ref{sec:aa} evaluate them by dynamic programming.

\section{Model parameters as data}\label{sec:tables}

\texttt{vdjtools} stores a model as a directory of long-format \texttt{polars} parquet tables plus a
\texttt{manifest.json} that names each event and lists its conditioning parents --- the edges of the
Bayesian network of Eq.~\eqref{eq:pvdj}. Changing a locus's conditioning (e.g.\ $P(D\mid J)$ versus a
richer $P(D\mid V,J)$) is therefore a one-line manifest edit, not a format change, and a $VJ$ locus is
simply the same schema with the $D$-tables and second insertion absent. Each table holds the
distribution of one event, normalized within its conditioning key by a single \texttt{polars}
\texttt{group\_by} --- which is exactly the M-step of \S\ref{sec:em}. Two conventions matter for the
$\Pgen$ recursions.

\paragraph{Palindromic deletions} A negative deletion count encodes templated palindromic (P-)
nucleotides. Each germline cut segment is pre-extended by up to \texttt{maxpal} reverse-complement
nucleotides; a ``deletion'' of $-k$ means $k$ P-nucleotides are retained. The stored count $\del$ is thus
\emph{biological} (negative $=$ P-nucleotides), while the dense array index used at run time is
$\del+\texttt{maxpal}\ge 0$.

\paragraph{Germline as a single source of truth} All V/D/J germline sequences and the CDR3 anchor
offsets resolve from \texttt{arda} by allele name; \texttt{arda}'s anchor convention is byte-identical to
\texttt{OLGA}'s (a $0$-based Cys\,104 / [FW]\,118 offset into the full germline), so no coordinate
conversion is needed. A model never mixes germline sources: bootstrap models imported from \texttt{OLGA}
keep \texttt{OLGA} germline to preserve exact-$\Pgen$ fidelity, whereas natively-inferred models use
\texttt{arda}.

\paragraph{The D-D extension} The schema is built to carry tandem-$D$ ($n_D\!\in\!\{0,1,2\}$)
rearrangements --- a $D$-count event, a second $D$-gene table and a $DD$ insertion --- which
\texttt{OLGA}/IGoR cannot represent. The bootstrap data used here (\texttt{OLGA}-drawn out-of-frame
sequences) contains no tandem $D$ by construction, so $D$-$D$ is exercised only structurally until real
data lands; the recursions below are written for $n_D\!\le\!1$ and degrade cleanly.

\section{Nucleotide generation probability}\label{sec:nt}

For a nucleotide query $\sigma$ the inner sum of Eq.~\eqref{eq:pgen} is a direct enumeration, because the
germline segments and both junctions are pinned by $\sigma$. Writing $N=|\sigma|$ and letting a $V$
(resp.\ $J$) contribute $\ell_V\ge1$ (resp.\ $\ell_J\ge1$) nucleotides --- a valid rearrangement is never
fully $V$- or $J$-deleted, matching \texttt{OLGA} ---
\begin{equation}\label{eq:pgennt}
  \Pgen(\sigma)=\!\!\sum_{V,J}\! P(V)P(J\!\mid\! V)\!\!
    \sum_{\substack{\ell_V,\ell_J\ge 1\\ \sigma[0{:}\ell_V]=\hat V,\ \sigma[N-\ell_J{:}]=\hat J}}
    \!\! P(\del_V)P(\del_J)\;\Xi\bigl(\sigma[\ell_V{:}N-\ell_J]\bigr),
\end{equation}
where for a $VJ$ locus $\Xi(\mu)=P_{\ins}(\mu)$ is the single-junction insertion weight of
Eq.~\eqref{eq:ins}, and for a $VDJ$ locus $\Xi$ is itself a sum over the $D$ gene, its two deletions and
its placement,
\begin{equation}\label{eq:dmiddle}
  \Xi(\mu)=\sum_D P(D\!\mid\! J)\!\!\sum_{\del_D^{5'},\del_D^{3'}}\!\! P(\del_D^{5'},\del_D^{3'})
    \sum_{\substack{\text{placements }p\\ \mu[p{:}p+\ell_D]=\hat D}}\!\!
    P_{\ins}\bigl(\mu[0{:}p]\bigr)\,P_{\ins}\bigl(\mu[p+\ell_D{:}]\bigr).
\end{equation}
Common-prefix / common-suffix matching restricts $\ell_V,\ell_J$ to the genes that can actually flank
$\sigma$, so Eq.~\eqref{eq:pgennt} is fast in practice. This nucleotide sum is the quantity the E-step
needs (\S\ref{sec:em}); the native port reproduces \texttt{OLGA}'s \texttt{compute\_nt\_CDR3\_pgen} to
machine precision.

\section{Amino-acid generation probability by transfer matrix}\label{sec:aa}

The amino-acid case is harder: the codon marginalization in Eq.~\eqref{eq:pgen} reintroduces a sum over
the unobserved junction nucleotides. Enumerating scenarios and summing synonymous codons position by
position (a codon-constrained dynamic program of width $|\Sigma|$ over the nucleotide state) is correct
but costs one $O(N)$ pass per $(D,\del_D^{5'},\del_D^{3'},\text{placement})$ skeleton --- quadratic in
$N$ and prohibitive for long CDR3s. The transfer-matrix method removes the placement and insertion sums.

\paragraph{Setup} Let $\alpha=\alpha_0\cdots\alpha_{L-1}$ be the query, $N=3L$, and $\sigma_i$ the (summed
over) nucleotide at position $i$. Codon $c$ imposes
$\tau(\sigma_{3c},\sigma_{3c+1},\sigma_{3c+2})=\alpha_c$. Because the codon and the insertion Markov chain
each look back at most two and one nucleotide respectively, a left-to-right dynamic program with state
$x_i=(\sigma_{i-1},\sigma_{i-2})$ suffices: it can (a) close a codon when $i\equiv 2\pmod 3$ and (b) carry
the Markov dependency. The width is $|\Sigma|=25$ (four nucleotides plus a start sentinel, squared). This
is the codon-marginalizing DP; the transfer matrix is what lets us reuse its partial results across all
scenarios.

\paragraph{The key structural fact} In the block layout of Eq.~\eqref{eq:concat},
$\hat V$--$\mathbf{m}^{VD}$--$\hat D$--$\mathbf{m}^{DJ}$--$\hat J$, the blocks are coupled only through
codon translation, never through an insertion Markov chain.

\begin{proposition}[Codon-only coupling]\label{prop:coupling}
Across every block boundary in Eq.~\eqref{eq:concat} the recombination weight $P(r)$ factors into a left
factor, the indicator(s) of the codon(s) straddling the boundary, and a right factor. No transition
$R(\cdot\mid\cdot)$ spans a boundary.
\end{proposition}
\begin{proof}[Sketch]
The $VD$ chain's leftmost nucleotide $m^{VD}_1$ is drawn from the bias $b_{VD}$ independent of the last
$V$ nucleotide, and its rightmost nucleotide carries no transition into $\hat D$; the $D$ block is
germline (fixed nucleotides, no chain); the $DJ$ chain, read $3'\!\to\!5'$, draws its $J$-proximal
nucleotide from $b_{DJ}$ independent of the first $J$ nucleotide, and its $D$-proximal nucleotide carries
no transition into $\hat D$. Hence every factor in Eq.~\eqref{eq:ins} lives strictly inside one block.
The only inter-block constraints are the codon indicators $\indic[\tau(\cdots)=\alpha_c]$ for codons that
straddle a boundary. \qed
\end{proof}

Proposition~\ref{prop:coupling} licenses computing the left and right halves of the sum \emph{once each}
and stitching them, instead of recomputing a full pass per scenario.

\paragraph{Left partial sums} Define, for each boundary position $p$ (the position at which $\hat D$
starts) and each nucleotide state $x=(\sigma_{p-1},\sigma_{p-2})$,
\begin{equation}\label{eq:lambda}
  \Lambda_p(x)=\!\!\sum_{\substack{V,\,\del_V,\,\ell_{VD}\\ \ell_V+\ell_{VD}=p}}\!\!
    P(V)P(\del_V)\,P_{\ins}(\ell_{VD})\;\bigl[\text{$VD$ Markov}\bigr]\;
    \prod_{\text{codons }\subset[0,p)}\!\!\indic[\tau=\alpha],
\end{equation}
the total weight of all ways the $V$ germline (with $\ell_V\ge1$) plus the $VD$ insertion fill
$[0,p)$ ending in state $x$, with every codon that lies wholly in $[0,p)$ satisfied. $\Lambda$ is built
by one forward recursion: germline seeds are deposited at $p=\ell_V$, then the $VD$ transfer step
--- multiply by $b_{VD}$ on the first inserted nucleotide and $R_{VD}$ thereafter, close codons at
$p\equiv2\pmod3$, and add $P_{\ins}(\ell_{VD})$ at each length --- extends them, in $O(N|\Sigma|)$ total.

\paragraph{Right partial sums} Symmetrically, for each $D$ gene and boundary $p$ (where $\hat D$ ends),
\begin{equation}\label{eq:rho}
  \Psi^D_p(y)=\!\!\sum_{J,\,\del_J,\,\ell_{DJ}}\!\!
    P(D\!\mid\! J)P(J)P(\del_J)\,P_{\ins}(\ell_{DJ})\;\bigl[\text{$DJ$ Markov}\bigr]\;
    \prod_{\text{codons }\subset[p,N)}\!\!\indic[\tau=\alpha],
\end{equation}
with $y=(\sigma_p,\sigma_{p+1})$. Because the $DJ$ chain runs $3'\!\to\!5'$, $\Psi^D$ is built by a
\emph{backward} recursion from the $J$ boundary --- one per $D$ gene, the $P(D\mid J)$ weight folded into
the seed. There are at most $n_D$ such passes.

\paragraph{Stitching} Fixing $D$, its deletions (hence the fixed germline string $\hat D$ of length
$\ell_D$) and its start $p_0$, thread the germline nucleotides of $\hat D$ through the DP from $\Lambda_{p_0}$
to a state distribution at $p_1=p_0+\ell_D$ (closing any codons interior to $\hat D$, no Markov weight),
then contract with $\Psi^D_{p_1}$. By Proposition~\ref{prop:coupling} at most one codon straddles the
$D/DJ$ boundary $p_1$; the contraction is a $4\times4$ (or scalar, when $p_1\equiv0$) sum over that
codon's free nucleotides. Summing over $D$, its deletions and $p_0$,
\begin{equation}\label{eq:stitch}
  \Pgen(\alpha)=\sum_D\ \sum_{\del_D^{5'},\del_D^{3'}} P(\del_D^{5'},\del_D^{3'})
    \sum_{p_0}\ \Bigl\langle\, \mathcal{T}_{\hat D}\,\Lambda_{p_0}\,,\ \Psi^D_{p_1}\,\Bigr\rangle_{p_1},
\end{equation}
where $\mathcal{T}_{\hat D}$ is the germline-threading operator and $\langle\cdot,\cdot\rangle_{p_1}$ the
codon-contracting stitch at $p_1$. The insertion-length sums are already inside $\Lambda$ and $\Psi^D$, so
the outer work is $O\!\bigl(n_D\cdot\#\!\del_D\cdot N\cdot\ell_D\bigr)$ with $\ell_D\ll N$ --- versus the
enumeration's extra factor of $N$ from re-running the DP per placement. For a $VJ$ locus there is no $D$
and no split: a single codon-marginalizing DP over $V$-germline\,--\,$VJ$-insertion\,--\,$J$-germline
templates suffices, which is already inexpensive.

\begin{remark}
This is the $\Pi_L\!\cdot\!\Pi_R$ factorization of \cite{Murugan2012,Sethna2019} expressed on the
\texttt{polars} tables. Where \texttt{OLGA} handles the $D$ block with a sliding right-deletion recursion
folded into $\Pi_R$, we keep $\Psi^D$ purely $J$/$DJ$ and slide a short explicit germline bridge
(Eq.~\eqref{eq:stitch}); the two agree to machine precision (\S\ref{sec:bench}), and the explicit bridge
is what generalizes to the tandem-$D$ event.
\end{remark}

\section{Correctness and benchmark}\label{sec:bench}

The native transfer matrix is validated against two independent oracles: the transparent
amino-acid scenario enumeration of \S\ref{sec:aa} (pure Python), and \texttt{OLGA} itself
\cite{Sethna2019}. On the human TRB model it reproduces the published worked example
$\Pgen(\texttt{CAWSVAPDRGGYTF}\mid \mathrm{TRBV30},\mathrm{TRBJ1\text{-}2})=1.2036\times10^{-10}$ and agrees
with \texttt{OLGA} to a maximum relative error of $1.2\times10^{-15}$ over $25$ marginalized (all V/J/D)
and $6$ V/J-restricted generated CDR3s --- i.e.\ to machine precision. Table~\ref{tab:bench} reports
timing, measured by \texttt{appendix/bench\_pgen.py} on the human TRB \texttt{OLGA} default model,
single-threaded, over $20$ marginalized queries.

\begin{table}[htbp]\centering
\caption{Amino-acid $\Pgen$ throughput on human TRB (single thread, marginalized over all V/J/D).
\textsc{tm} is the Eq.~\eqref{eq:stitch} transfer matrix in the native \texttt{\_core}; \textsc{olga} is
\texttt{compute\_aa\_CDR3\_pgen}. Lower is better.}\label{tab:bench}
\footnotesize
\begin{tabular}{lrrr}\toprule
Method & ms / sequence & vs \textsc{olga} & max.\ rel.\ error vs \textsc{olga} \\\midrule
\textsc{olga}               & $5.3$          & $1.0\times$                   & --- \\
\textsc{tm} (native \texttt{\_core}) & $\mathbf{0.64}$ & $\mathbf{8.3\times}$ & $1.2\times10^{-15}$ \\
\bottomrule\end{tabular}
\end{table}

\noindent The pure-Python scenario enumeration of \S\ref{sec:aa} is exact and serves as the correctness
oracle, but is not a runtime path: unrestricted, it costs on the order of minutes per query (its extra
factor of $N$ and its lack of the left/right reuse), which is exactly the cost the transfer matrix
removes.

\section{Degenerate, agnostic and Hamming-ball queries}\label{sec:motif}

Three capabilities fall out of the same transfer matrix at essentially no extra cost. First, the
codon test at each amino-acid position is a \emph{membership test against an allowed-codon set} (a
$64$-bit mask), not an equality against one residue; a wildcard mask (any of the $20$ amino acids)
or any degenerate set is therefore evaluated in a single pass, giving motif/regex $\Pgen$ without
enumerating the matching sequences. Second, conditioning on a $V$ and/or $J$ allele is a usage mask
over the gene sums of Eqs.~\eqref{eq:lambda},~\eqref{eq:rho}; the empty restriction marginalizes over
all functional genes --- \emph{V-agnostic} / \emph{J-agnostic} $\Pgen$.

Combining the two gives a fast one-mismatch query. The total $\Pgen$ of an amino-acid CDR3 and every
sequence within Hamming distance $1$ (one substitution) is, by inclusion--exclusion,
\begin{equation}\label{eq:ham1}
  \Pgen^{\le 1}(\alpha)=\sum_{k=0}^{L-1}\Pgen(\alpha_{k\to *})-(L-1)\,\Pgen(\alpha),
\end{equation}
where $\alpha_{k\to *}$ wildcards position $k$. \texttt{OLGA} uses the same identity but pays a full
(slow) transfer-matrix pass for each of the $L{+}1$ terms; here each term is one fast pass
(Eq.~\eqref{eq:stitch}) with a single wildcard mask, so the native \texttt{\_core} evaluates
Eq.~\eqref{eq:ham1} at $8.7\times$ \texttt{OLGA}'s throughput on human TRB ($67.9$ vs $7.8$~ms/query)
and $2.5\times$ on human TRA ($25.1$ vs $10.1$~ms/query), matching it to a relative error of
$10^{-15}$ throughout. A codon-boundary forward/backward sweep would collapse the $L{+}1$ passes into
a single sweep for $VJ$ loci; the $D$-placement sum couples positions in $VDJ$ loci, so there the
$L{+}1$-pass identity is retained.

\section{Generation and inference on the same tables}\label{sec:em}

\paragraph{Generation} Ancestral sampling walks the Bayesian network of Eq.~\eqref{eq:pvdj} top-down:
draw $(V,J[,D])$, the deletions, the insertion lengths, then the insertion nucleotides from the Markov
chain, and assemble Eq.~\eqref{eq:concat}. Restricting to functional genes and enforcing $\ell_V,\ell_J\ge1$
keeps every draw scoreable; sampled V/J-usage and CDR3-length distributions match the model marginals.

\paragraph{Inference} Model parameters are estimated from out-of-frame reads by
expectation--maximization. The E-step is exactly the scenario sum of \S\ref{sec:nt}: for each read it
computes the posterior weight $P(r\mid\sigma)=P(r)/\Pgen(\sigma)$ of every compatible scenario and
accumulates soft counts into arrays laid out like the model tables; optionally the enumeration is
restricted to the genes \texttt{arda} aligns (an \emph{aligned-gene mask}), which prunes the V/D/J sums
without changing the fixed point. The M-step renormalizes each event table within its conditioning key ---
a single \texttt{polars} \texttt{group\_by} --- reproducing the manifest's Bayesian network. On
\texttt{OLGA}-drawn data this is a closed-loop oracle: EM must recover the generating marginals, giving a
self-checking target before any real data exists. The E-step inner loop is the second native hot path (the
first being $\Pgen$); with the aligned-gene mask it makes $VDJ$ inference practical.

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}
